---
title: "Weather and Corn Yield Regressions"
author: "Derek Fedak"
date: "`r format(Sys.time(), '%A %d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
require(gridExtra)
#https://fedakd.github.io/weather_corn_regressions/assignment.html
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)
```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)
```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")
```

### Download NASS corn yield data

```{r yield download}

# set our API key with NASS
nassqs_auth(key = "DFB1C7BB-A731-3630-8012-415D0AD83D4F")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a

Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r}
winnecorn <- cornyields[cornyields$county_ansi == 191,]

ggplot(winnecorn, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Corn yield (Bu/Acre)") +
  ggtitle("Corn yield over time, Winneshiek County") +
  geom_smooth(method = lm)
```

### Answer 1a

There is a positive correlation between corn `yield` and `year` in Winneshiek County, which illustrates that corn yield at a Bu/acre metric has increased over time from 1981 to 2021.

### Question 1b

Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r}
winnecorn$yearsq <- winnecorn$year^2

lm_cornyieldquad <- lm(yield ~ year + yearsq, winnecorn)
summary(lm_cornyieldquad)
winnecorn$fitted <- lm_cornyieldquad$fitted.values

ggplot(winnecorn) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  ggtitle("Corn yield over time and quadratic time model, Winneshiek County") +
  labs(x = "Year", y = "Corn yield (Bu/Acre)")
```

### Answer 1b

There is not evidence of a slowing yield growth in our quadratic time trend. Our `year` variable has a negative intercept while our `years^2` variable has a positive intercept. This would suggest a potential increasing `yield` growth over time. However, both time variables in our model are insignificant (>0.7, in this case), so we cannot conclusively say the yield growth rate is decreasing or increasing over time.


### Question 2 Time Series

Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r}
winnecornsummer <- inner_join(winnesummer, winnecorn, by="year") 

#Analyzing with just yield and Tmax


ggplot(winnecornsummer, mapping = aes(x = meantmax, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Mean summer max T (°C)", y = "Corn yield (Bu/Acre)") +
  ggtitle("Mean summer max T on corn yield, Winneshiek County") +
  geom_smooth(method = lm)

#Linear Models
#Linear model of just yield and mean Tmax
lm_cornsummertmax <- lm(yield ~ meantmax, winnecornsummer)
summary(lm_cornsummertmax)
#Linear model of yield and mean Tmax + year
lm_cornsummertmax1 <- lm(yield ~ year + meantmax, winnecornsummer)
summary(lm_cornsummertmax1)

#Quadratic Models
#Quadratic model using year, year^2, mean Tmax, and mean Tmax^2
winnecornsummer$meantxmaxsq <- winnecornsummer$meantmax^2
lm_cornyieldsummertquad <- lm(yield ~ year + yearsq + meantmax + meantxmaxsq, winnecornsummer)
summary(lm_cornyieldsummertquad)
winnecornsummer$fitted <- lm_cornyieldsummertquad$fitted.values
#Quadratic model using year, mean Tmax, and mean Tmax^2
lm_cornyieldsummertquad2 <- lm(yield ~ year + meantmax + meantxmaxsq, winnecornsummer)
summary(lm_cornyieldsummertquad2)
winnecornsummer$fitted2 <- lm_cornyieldsummertquad2$fitted.values

#AIC
AIC(lm_cornsummertmax, lm_cornsummertmax1, lm_cornyieldsummertquad, lm_cornyieldsummertquad2)

p1 <- ggplot(winnecornsummer) +
      geom_point(mapping = aes(x = year, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = year, y = fitted2, col="Fitted")) +
      theme_bw() +
      ggtitle("Corn yield over time, quadratic time model, Winneshiek County") +
      labs(x = "Year", y = "Corn yield (Bu/Acre)")
  
p2 <- ggplot(winnecornsummer) +
      geom_point(mapping = aes(x = meantmax, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = meantmax, y = fitted2, col="Fitted")) +
      theme_bw() +
      ggtitle("Corn yield over summer max temperature, and quadratic temperature model, Winneshiek County") +
      labs(x = "Mean summer max T (°C) squared", y = "Corn yield (Bu/Acre)")
grid.arrange(p1, p2)
```

### Answer 2

Our initial linear model of `mean summer max temperature` on corn `yield` shows that our slope is weak and insignificant. This makes intuitive sense since plants have an optimal temperature growth range, which appears to be 25-28 °C. Outside of that range, yields decrease. When we include `year` as a predictor variable, it does become significant (p<0.001). `Year` alone was a poor predictor (Answer 1a) but becomes significant when analyzed with `mean max summer temperature`, which could relate to other environmental variables like drought, fire, winter temperature, or flood for a given year.

In our quadratic model using `year`, `year^2`, `mean summer max temperature`, and `mean summer max temperature^2`, both temperature predictors are significant (p<0.001), while the year predictors are insignificant. In the end, our best model used `year`, `mean summer max temperature`, and `mean summer max temperature^2` as predictors were all highly significant (p<0.001) with an AIC of 315.63. This suggests that an optimal temperature range modeled with extreme environmental disturbance years best predicts a year's corn yield.

### Question 3 Cross-Section

Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r}
corn2018 <- cornyields[cornyields$year == 2018,]
corn2018$countyfp <- as.factor(corn2018$county_ansi)

summer2018 <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  filter(year == 2018) %>%
  group_by(year, countyfp) %>%
  summarize(meantmaxsum = mean(tmax))

#Just using first 59 days in 2018 and not end of 2017
winter2018 <- tmaxdf %>%
  filter((doy <= 59) & !is.na(tmax)) %>%
  filter(year == 2018) %>%
  group_by(year, countyfp) %>%
  summarize(meantmaxwin = mean(tmax))

corn2018 <- left_join(corn2018, summer2018, keep=F, by="countyfp")
corn2018 <- left_join(corn2018, winter2018, keep=F, by="countyfp")
corn2018$meantmaxsumsq <- corn2018$meantmaxsum^2
corn2018$meantmaxwinsq <- corn2018$meantmaxwin^2

#Linear models analyzing summer, winter, and both combined
lm_2018cornsummer <- lm(yield ~ meantmaxsum, corn2018)
summary(lm_2018cornsummer)
lm_2018cornwinter <- lm(yield ~ meantmaxwin, corn2018)
summary(lm_2018cornwinter)
lm_2018cornall <- lm(yield ~ meantmaxsum + meantmaxwin, corn2018)
summary(lm_2018cornall)
#Quadratic models analyzing summer, winter, and both combined
lm_2018cornsummer2 <- lm(yield ~ meantmaxsum + meantmaxsumsq, corn2018)
summary(lm_2018cornsummer2)
lm_2018cornwinter2 <- lm(yield ~ meantmaxwin + meantmaxwinsq, corn2018)
summary(lm_2018cornwinter2)
lm_2018cornall2 <- lm(yield ~ meantmaxsum + meantmaxwin + meantmaxsumsq + meantmaxwinsq, corn2018)
summary(lm_2018cornall2)
AIC(lm_2018cornsummer, lm_2018cornsummer2, lm_2018cornwinter, lm_2018cornwinter2, lm_2018cornall, lm_2018cornall2)

p1 <- ggplot(corn2018, mapping = aes(x = meantmaxsum, y = yield)) +
      geom_point() +
      theme_bw() +
      labs(x = "Mean summer max T (°C)", y = "Corn yield (Bu/Acre)") +
      ggtitle("Mean summer max T on corn yield across Iowa, 2018") +
      geom_smooth(method = lm)

p2 <- ggplot(corn2018, mapping = aes(x = meantmaxwin, y = yield)) +
      geom_point() +
      theme_bw() +
      labs(x = "Mean winter max T (°C)", y = "Corn yield (Bu/Acre)") +
      ggtitle("Mean winter max T on corn yield across Iowa, 2018") +
      geom_smooth(method = lm)
grid.arrange(p1, p2)
```

### Answer 3

`Mean max summer temperature` modeled across all Iowa counties as a predictor for corn `yield` in 2018 is a weak but appropriate approach (p=0.06). Using a similar model with `mean max winter temperature` for the first 59 days of 2018 was a worse model (p=0.9). I ran several combinations using both summer and winter temperature as well as their squares in a quadratic approach, and summmer temperature remained the best predictor but only significant in a quadratic model (p<0.01). I believe this is due to corn's optimal temperature range being 25-28 °C. The best AIC value *793.56) was from a quadratic model using both winter and summer temperatures and their squares.

### Question 4 Panel

One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r}
cornyields$countyfp <- as.factor(cornyields$county_ansi)
allsummer <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  summarize(meantmax = mean(tmax))

allcorn <- left_join(cornyields, allsummer, by=c("countyfp", "year")) %>%
  na.omit(.)
allcorn$meantmaxsq <- allcorn$meantmax^2

lm_allcorn <- lm(yield ~ countyfp + meantmax + meantmaxsq, allcorn)
summary(lm_allcorn)
AIC(lm_allcorn)
allcorn$fitted <- lm_allcorn$fitted.values

p1 <- ggplot(allcorn) +
      geom_point(mapping = aes(x = meantmax, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = meantmax, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Mean max summer temperature (°C)", y = "Corn yield (Bu/Acre)") +
      ggtitle("Actual vs. fitted corn yields over temperature, Iowa")

p2 <- ggplot(allcorn) +
      geom_point(mapping = aes(x = year, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = year, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Year", y = "Corn yield (Bu/Acre)") +
      ggtitle("Actual vs. fitted corn yields over time, Iowa")
grid.arrange(p1, p2)
```

### Answer 4

Modeling county as a factor across all years in a panel regression does make our temperature variables (`mean summer max T` and `mean summer max T^2`) highly significant (p<0.001). It does drastically increase our AIC to 36,402 due to the increased number of predictor variables. This suggests that over the entire study area, `mean max summer temperature` is a much better predictor of corn `yield` than the individual `county`.

### Question 5 Soybeans

Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r}
# parameters to query on 
params <- list(commodity_desc = "SOYBEANS", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soyyieldsall <- nassqs_yields(params)

soyyieldsall$county_ansi <- as.numeric(soyyieldsall$county_ansi)
soyyieldsall$yield <- as.numeric(soyyieldsall$Value)

# clean and filter this dataset
soyyields <- select(soyyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyyields <- tibble(cornyields)

#Time-series analysis of Buena Vista County
bvsummer <- tmaxdf %>%
  filter(countyfp==21 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))
bvwinter <- tmaxdf %>%
  filter(countyfp==21 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

bvsoy <- soyyields[soyyields$county_ansi == 21,]

bvsoysummer <- inner_join(bvsummer, bvsoy, by="year") 

#Visualization

p1 <- ggplot(bvsoysummer, mapping = aes(x = year, y = yield)) +
      geom_point() +
      theme_bw() +
      labs(x = "Year", y = "Soybean yield (Bu/Acre)") +
      ggtitle("Soybean yield over time, Buena Vista County") +
      geom_smooth(method = lm)

p2 <- ggplot(bvsoysummer, mapping = aes(x = meantmax, y = yield)) +
      geom_point() +
      theme_bw() +
      labs(x = "Mean summer max T (°C)", y = "Soybean yield (Bu/Acre)") +
      ggtitle("Mean summer max T on soybean yield, Buena Vista County") +
      geom_smooth(method = lm)
grid.arrange(p1,p2)

#Linear Models
#Linear model of just yield and mean Tmax
lm_bvsummertmax <- lm(yield ~ meantmax, bvsoysummer)
summary(lm_bvsummertmax)
#Linear model of yield and mean Tmax + year
lm_bvsummertmax1 <- lm(yield ~ year + meantmax, bvsoysummer)
summary(lm_bvsummertmax1)

#Quadratic Models
#Quadratic model using year, year^2, mean Tmax, and mean Tmax^2
bvsoysummer$yearsq <- bvsoysummer$year^2
bvsoysummer$meantxmaxsq <- bvsoysummer$meantmax^2
lm_soyyieldsummertquad <- lm(yield ~ year + yearsq + meantmax + meantxmaxsq, bvsoysummer)
summary(lm_soyyieldsummertquad)
bvsoysummer$fitted <- lm_soyyieldsummertquad$fitted.values
#Quadratic model using year, mean Tmax, and mean Tmax^2
lm_soyyieldsummertquad2 <- lm(yield ~ year + meantmax + meantxmaxsq, bvsoysummer)
summary(lm_soyyieldsummertquad2)
bvsoysummer$fitted2 <- lm_soyyieldsummertquad2$fitted.values

#AIC
AIC(lm_bvsummertmax, lm_bvsummertmax1, lm_soyyieldsummertquad, lm_soyyieldsummertquad2)

p1 <- ggplot(bvsoysummer) +
      geom_point(mapping = aes(x = year, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = year, y = fitted2, col="Fitted")) +
      theme_bw() +
      ggtitle("Soy yield over time, quadratic time model, Buena Vista County") +
      labs(x = "Year", y = "Soy yield (Bu/Acre)")
  
p2 <- ggplot(bvsoysummer) +
      geom_point(mapping = aes(x = meantmax, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = meantmax, y = fitted2, col="Fitted")) +
      theme_bw() +
      ggtitle("Soy yield over summer max temperature, and quadratic temperature model, Buena Vista County") +
      labs(x = "Mean summer max T (°C) squared", y = "Soy yield (Bu/Acre)")
grid.arrange(p1, p2)
```

### Answer 5

Blah

### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
